Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-22 threshold/draft.Rnw
Added experimental/prototype implementation; a few typos in the paper

XXX add discussion of applications (especially genomics?) and previous work,
then on to the preliminaries:
\[\]

@@

Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $U^TU = I$, $V\in\mathcal{R}^{n\times k}$,
$V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$ to be
the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$. The span
of the columns of $V$ form an orthonormal basis of the range of the correlation
matrix $A^TA$.

@@

\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{cordist}
\mathrm{cor}(a_i,a_j) = a_i^Ta_j = 1 - \|a_i - a_j\|^2/2.
\end{equation}
In particular, for a given correlation threshold $t$, $0<t<1$, pairs
of vectors $a_i$ and $a_j$ such that $\|a_i - a_j\|^2 > 2(1-t)$ lie
outside the correlation threshold.
\end{lemma}
\begin{proof} (XXX perhaps omit this)
\begin{eqnarray*}
\|a_i - a_j\|^2 &=& (a_i - a_j)^T(a_i - a_j) \\
&=& a_i^T a_i + a_j^T a_j - 2a_i^T a_j \\
&=& \|a_i\|^2 + \|a_j\|^2 - 2a_i^T a_j \\
&=& 2 - 2a_i^T a_j.
\end{eqnarray*}
The lemma follows by solving for $a_i^Ta_j$.
\end{proof}

@@

\begin{figure}[!ht]
\label{fig1}
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(BicatYeast)
A=t(BicatYeast)
s=svd(A)
i=order(s$v[,1])
plot(s$v[i,1], ylab=expression(Pi * v[1]), xlab="Sorted index", pch='.')
t=0.99
abline(h=c(-0.05,-0.05+sqrt(2*(1-t))/s$d[1]),col=2,lwd=0.5)
@
\caption{Example plot of the ordered entries of $v_1$ for the BicatYeast example from Section \ref{examples}. The red lines illustrate the interval corresponding threshold $t=0.99$. Points farther apart than this interval correspond to column vectors that do not meet the correlation threshold.}
\end{center}
\end{figure}

Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
difference matrix with diagonal entries $-1$, $i$th super-diagonal entries $1$,
and zeros elsewhere.  Consider just adjacent points of $P v_1$.  Then $D^1P
v_1$ consists of the differences of adjacent entries in $P v_1$.  Entries
such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$, where the exponentiation is
applied element-wise, correspond to pairs of vectors that do not meet
the correlation threshold. This is unlikely to rule out many pairs of vectors
in most problems. For example, the minimum distance between the points shown in
the example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma$, and no
pruning at all occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
pairs of vectors outside the correlation threshold. For example, including
only one more term finds that only 116 out of 418 pairs of adjacent vectors
in the example shown in Figure \ref{fig1} satisfy
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 \le 2(1-t),
\]
where exponentiation is applied element-wise to the vectors.
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including five terms reduces the total to only 11 pairs of vectors.
In general, using $p\le k$ terms boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values,
and exponentiation is applied element-wise.

Pairs of vectors beyond the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=1,2,\ldots,\ell$ in the above expressions.
Following this procedure, we can produce a well-pruned candidate set of
possible vector pairs that meet a given correlation threshold using $\ell$
matrix vector products with matrices of order $n \times p$, where $p$ is chosen
in practice such that $p<<k$. The steps of this procedure are outlined in the
following algorithm:

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$, maximum
subspace dimension $p<<n$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where $V_{1:p} = [v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
zeros otherwise.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs stops decreasing.
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

Algorithm \ref{prune} by construction guarantees that no pruned vector pair
meets the given correlation threshold; that is, the pruning can not produce any
``false negatives.''  The algorithm typically prunes the vast majority of pairs
of vectors outside the correlation threshold with truncated singular value
decompositions of only relatively low dimension $p$.  The choice of $p$ is a
balance between work in computing a truncated SVD and work required by the
$\ell$ matrix vector products of order $n\times p$ in step 4.

@@

We have so far assumed that the mean of each column vector of $A$ is zero and
the columns are scaled to have norm 1, but we really want a method that works
with general matrices $A$. We've also implicitly assumed that we can
efficiently compute a relatively low-dimensional truncated SVD of potentially
large  matrices required by step 1 of Algorithm \ref{prune}.  Fortunately,
both needs are met by one algorithm.

The augmented, implicitly restarted Lanczos bidiagonalization (IRLBA) method of
Baglama and Reichel \cite{br} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
The R implementation includes efficient methods for re-starting the algorithm
to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
efficient methods for computing the SVD of scaled and centered matrices.

>>>>

XXX add discussion of applications (genomics?) and previous work,
then on to the preliminaries:
\[\]

@@

Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $U^TU = I$, $V\in\mathcal{R}^{n\times k}$,
$V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$ to be
the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$. The span
of the columns of $V$ form an orthonormal basis of the range of the correlation
matrix $A^TA$.

@@

\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{cordist}
\mathrm{cor}(a_i,a_j) = a_i^Ta_j = 1 - \|a_i - a_j\|^2/2.
\end{equation}
In particular, for a given correlation threshold $t$, $0<t<1$, pairs
of vectors $a_i$ and $a_j$ such that $\|a_i - a_j\|^2 > 2(1-t)$ lie
below the correlation threshold.
\end{lemma}
\begin{proof} (XXX perhaps omit this)
\begin{eqnarray*}
\|a_i - a_j\|^2 &=& (a_i - a_j)^T(a_i - a_j) \\
&=& a_i^T a_i + a_j^T a_j - 2a_i^T a_j \\
&=& \|a_i\|^2 + \|a_j\|^2 - 2a_i^T a_j \\
&=& 2 - 2a_i^T a_j.
\end{eqnarray*}
The lemma follows by solving for $a_i^Ta_j$.
\end{proof}

@@

\begin{figure}[!ht]
\label{fig1}
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(BicatYeast)
A=t(BicatYeast)
s=svd(A)
i=order(s$v[,1])
plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
t=0.99
abline(h=c(-0.05,-0.05+sqrt(2*(1-t))/s$d[1]),col=2,lwd=0.5)
@
\caption{Example plot of the ordered entries of $v_1$ for the BicatYeast example from Section \ref{examples}. The red lines illustrate the interval corresponding threshold $t=0.99$. Points farther apart than this interval correspond to column vectors that do not meet the correlation threshold.}
\end{center}
\end{figure}

Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
difference matrix with $-1$ along the diagonal, $1$ along the $i$th
super-diagonal, and zeros elsewhere.  Consider just adjacent points of $P v_1$.
Then $D^1P v_1$ consists of the differences of adjacent entries in $P v_1$.
Entries such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$
correspond to pairs of vectors that do not meet the correlation threshold,
(exponentiation is applied to the vector element-wise).
However, this observation is unlikely to rule out many pairs of vectors in
most problems. For example, the minimum distance between the points shown in
the example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma$, and no
pruning at all occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
pairs of vectors below the correlation threshold. For example, including
only one more term finds that 72\% of adjacent vectors fall below the
correlation threshold
in the example shown in Figure \ref{fig1} and are pruned by identifying
indices such that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t),
\]
where exponentiation is applied element-wise to the vectors.
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including five terms prunes over 97\% of adjacent vector pairs,
leaving only 11 candidate pairs of adjacent vectors.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values,
and exponentiation is applied to vectors element-wise.

Pairs of vectors below the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=1,2,\ldots,\ell$ in the above expressions.
Following this procedure, we can produce a well-pruned candidate set of
possible vector pairs that meet a given correlation threshold using $\ell$
matrix vector products with matrices of order $n \times p$, where $p$ is chosen
in practice such that $p<<k$. The steps of this procedure are outlined in the
following algorithm:

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$, maximum
subspace dimension $p<<n$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where $V_{1:p} = [v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
zeros otherwise.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

Algorithm \ref{prune} by construction guarantees that no pruned vector pair
meets the given correlation threshold; that is, the pruning can not produce any
``false negatives.''  The algorithm typically prunes the vast majority of pairs
of vectors below the correlation threshold with truncated singular value
decompositions of only relatively low dimension $p$.  The choice of $p$ is a
balance between work in computing a truncated SVD and work required by the
$\ell$ matrix vector products of order $n\times p$ in step 4.

@@

We have so far assumed that the mean of each column vector of $A$ is zero and
the columns are scaled to have norm 1, but we really want a method that works
with general matrices $A$. We've also implicitly assumed that we can
efficiently compute a relatively low-rank truncated SVD of potentially
large  matrices required by step 1 of Algorithm \ref{prune}.  Fortunately,
both needs are met by one algorithm.

The augmented, implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
of Baglama and Reichel \cite{br} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
The R implementation includes efficient methods for re-starting the algorithm
to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
efficient methods for computing the SVD of scaled and centered matrices without
explicitly forming a new matrix.
